Letter: Performance models for \\ high-resolution ice sheet simulations

### 2022-06-23 paper/perfmod.tex
recent ism refs, and start on intro

\abstract{\today. FIXME}

@@

FIXME INCREASING UNDERSTANDING OF TIME STEP LIMITS \citep{Chengetal2017,Robinsonetal2022}; MORE THAN SIA CASE \citep[for comparison]{HindmarshPayne1996}, POSSIBILITY OF IMPLICIT SOLVERS \citep{Bueler2016,WirbelJarosch2020}, NEED PERFORMANCE MODEL TO UNDERSTAND TRADE-OFFS

FIXME \citep{GreveBlatter2009,SchoofHewitt2013} for SIA and Stokes equations

FIXME We expect that time-stepping glacier and ice-sheet numerical models will become faster, and thus more useful tools for glaciologists, when multigrid is applied to the solution of the problem(s) solved at each time-step.  This hope can be understood and assessed by a \emph{performance model} for the scheme, one which combines computational work estimates for subproblems into an asymptotic, in the limit of highly-refined meshes, description of of the work required for a given simulation task.

@@

For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ where $D$ is a representative effective diffusivity in $\text{km}^2\,\text{a}^{-1}$ \citep{Bueleretal2005,HindmarshPayne1996}.\footnote{Recall $\displaystyle D = \frac{2 A (\rho g)^n}{n+2} H^{n+2} |\grad s|^{n-1}$ where $A$ is ice softness, $\rho$ is ice density, $g$ is gravity, $H$ is ice thickness, and $s$ is ice surface elevation.}
For Stokes dynamics the question of explicit time-stepping stability is largely unexplored.  We propose a \emph{pessimistic} explicit time-step stability paradigm, namely the above SIA restriction $\Delta t < O(D^{-1} \Delta x^2)$.  This is motivated by the small geometric aspect ratio analysis of the Stokes model leading to the SIA model \citep{GreveBlatter2009}.  Alternatively one might subscribe to an \emph{optimistic} paradigm supposing that stability is controlled by the advective restriction $\Delta t < O(U^{-1} \Delta x)$ for some typical horizontal velocity scale $U$ in $\text{km}\,\text{a}^{-1}$.

>>>>

\abstract{Modern numerical ice sheet models, with horizontal resolutions of a few kilometers or less, compute evolving ice geometry and velocity fields using various widely-discussed stress-balance approximations and boundary conditions.  The practical performance of such numerical models is, however, often dominated by the stability restrictions of currently-standard explicit time-stepping.  Such models take many short, stability-limited time steps because ice thickness is updated following each velocity solution.  We clarify the performance of numerical model choices, in the high-resolution limit, by quantifying computational effort in terms of major parameters.  The potential performance of coming generations of Stokes dynamics and/or implicit time-stepping numerical models is assessed.}

@@

Numerical ice sheet models with evolving ice geometry are now in common use for scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenland \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and quantification of long-term glacial erosion \citep{SeguinotDelaney2021}.  It is generally accepted that horizontal mesh or grid cells must be smaller than about 10 km in order to generate valid results.  Whether using local mesh refinement \citep[for example]{Hoffmanetal2018} or not, resolutions of a few kilometers to one kilometer \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019} have been used at ice sheet scale.  With available computational resources, sub-kilometer meshes will become increasingly common even for long-term, whole ice sheet scale studies.

The current generation of ice sheet models use a variety of stress balances, from simple approximations like the shallow ice approximation (SIA), through ``hybrid'' \citep[for example]{Winkelmannetal2011} and higher-order approximations \citep{GreveBlatter2009}, to the non-shallow Stokes approximation.  With very few exceptions, however, current models alternate between solving the stress balance for velocity, using geometry from the previous time step, and then updating the ice thickness and/or surface elevation.  The latter geometry-update operation, using the mass continuity or surface kinematical equations \cite{GreveBlatter2009}, is where interaction between the ice sheet and the climate is (primarily) simulated, especially through surface mass balance (SMB) and calving processes.  Such schemes are called \emph{explicit time-stepping} for the coupled mass and momentum system which describes ice dynamics.

For simpler partial differential equation problems, the conditional stability of explicit time-stepping schemes is well-understood \citep{LeVeque2007}.  The stability limitations of SIA numerical models have been understood, at least empirically, for some time \citep{HindmarshPayne1996}.  More recent studies have focussed on time-step limits applicable to models using hybrid, higher-order, and Stokes dynamics \citep{Chengetal2017,Robinsonetal2022} or on techniques to mitigate such limits \citep{LofgrenAhlkronaHelanow2021}.

FIXME POSSIBILITY OF IMPLICIT SOLVERS \citep{Bueler2016,WirbelJarosch2020}

FIXME see \citep{GreveBlatter2009,SchoofHewitt2013} for SIA and Stokes equations; CONFUSION FROM TWO FORMATS
    $$\frac{\partial H}{\partial t} + \Div \left(U H\right) = a$$
    $$\frac{\partial H}{\partial t} = \Div \left(D \grad s \right) + a$$
where FIXME; Recall $\displaystyle D = \frac{2 A (\rho g)^n}{n+2} H^{n+2} |\grad s|^{n-1}$ where $A$ is ice softness, $\rho$ is ice density, $g$ is gravity, $H$ is ice thickness, and $s$ is ice surface elevation; TSUNAMI WAVES EXAMPLE OF FORMER NOT LATER; NOTHING LIKE TSUNMAMIS IN ICE BECAUSE ICE FLOWS DOWNHILL WITH RARE EXCEPTIONS

FIXME SITUATION COMPLICATED AND PRONE TO MISUNDERSTANDING; NEED PERFORMANCE MODEL TO UNDERSTAND TRADE-OFFS; We expect that time-stepping glacier and ice-sheet numerical models will become faster, and thus more useful tools for glaciologists, when multigrid is applied to the solution of the problem(s) solved at each time-step.  This hope can be understood and assessed by a \emph{performance model} for the scheme, one which combines computational work estimates for subproblems into an asymptotic, in the limit of highly-refined meshes, description of of the work required for a given simulation task.

@@

For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ where $D$ is a representative effective diffusivity in $\text{km}^2\,\text{a}^{-1}$ \citep{Bueleretal2005,HindmarshPayne1996}.
For Stokes dynamics the question of explicit time-stepping stability is largely unexplored.  We propose a \emph{pessimistic} explicit time-step stability paradigm, namely the above SIA restriction $\Delta t < O(D^{-1} \Delta x^2)$.  This is motivated by the small geometric aspect ratio analysis of the Stokes model leading to the SIA model \citep{GreveBlatter2009}.  Alternatively one might subscribe to an \emph{optimistic} paradigm supposing that stability is controlled by the advective restriction $\Delta t < O(U^{-1} \Delta x)$ for some typical horizontal velocity scale $U$ in $\text{km}\,\text{a}^{-1}$.
